\chapter{Conclusion}

\section{Introduction}
This chapter summarises the key findings of the study, discusses theoretical and practical implications, and outlines limitations and directions for future research.

\section{Summary of Findings}
This dissertation examined how artificial intelligence (AI) and generative artificial intelligence (GenAI) influence consumer perception in marketing. The study argued that consumer perception of AI is not adequately explained by describing AI as either beneficial or harmful in general terms. Instead, the literature suggests that consumer responses are shaped by recurring tensions that emerge when AI offers practical value while simultaneously creating psychological, emotional, and ethical concerns.

Using a thematic literature review informed by Puntoni et al.'s experiential framework, the dissertation identified four overarching tensions: data capture, classification, delegation, and social interaction. These were interpreted through the consumer-perception tensions of served versus exploited, understood versus misunderstood, empowered versus replaced, and connected versus alienated. Taken together, these tensions show that consumer responses to AI in marketing are conditional rather than uniform.

The first key finding is that AI tends to generate more favourable responses when data use is perceived as proportionate, transparent, and valuable. Consumers are more likely to accept AI when it improves relevance, convenience, and efficiency without making them feel monitored or exploited. The second key finding is that personalisation does not automatically create a sense of individuality. AI can make consumers feel understood, but it can also make them feel reduced to a category when recommendations seem generic, stereotypical, or insufficiently responsive to personal context.

The third key finding is that delegation produces both relief and resistance. AI is often welcomed when it reduces effort and supports decision-making, but it becomes more problematic when consumers feel that control, agency, or self-efficacy is being displaced. The fourth key finding is that authenticity becomes especially important when AI moves from back-end optimisation into visible content generation and social interaction. Consumers may accept AI in factual or utilitarian contexts, yet become more resistant in emotional, hedonic, or morally sensitive contexts where they expect human sincerity, empathy, or creative intention.

Across all four tensions, trust emerged as a cross-cutting mechanism rather than a standalone theme. Trust increases when AI is perceived as useful, legitimate, and contextually appropriate. It declines when AI appears opaque, manipulative, overly visible, or emotionally inauthentic. Behavioural outcomes such as engagement and purchase intention are therefore best understood as consequences of how these tensions are resolved in specific contexts.

\section{Theoretical Implications}
This study contributes to the literature in several ways. First, it offers a more precise conceptualisation of consumer perception in AI-mediated marketing by treating perception as an interpretive process that combines cognitive evaluation, affective response, and behavioural consequence. This helps move the discussion beyond narrower adoption-oriented approaches that are less suited to explaining visible and socially meaningful AI encounters in marketing.

Second, the dissertation contributes by showing that many apparent contradictions in the literature are more accurately understood as trade-offs or conditional effects. This matters theoretically because it shifts the focus from asking whether AI is positively or negatively perceived to asking under what conditions certain benefits are accepted and under what conditions those same benefits are reinterpreted as risks.

Third, the study demonstrates the value of Puntoni et al.'s experiential framework for organising the AI-consumer literature. By translating data capture, classification, delegation, and social interaction into four consumer-perception tensions, the dissertation provides a more integrated framework for connecting privacy, trust, authenticity, autonomy, and behavioural intention. This offers a clearer structure for future literature reviews and empirical research.

Finally, the dissertation extends existing discussion of GenAI by distinguishing between source authenticity and emotional authenticity. This distinction is theoretically useful because it clarifies that consumer resistance to AI-generated communication is not only about whether a message is AI-authored, but also about whether it feels emotionally sincere and contextually appropriate.

\section{Practical Implications}
From a practical perspective, the findings suggest that marketers should carefully balance the use of AI technologies with consumer expectations.

First, firms should not assume that technological sophistication automatically improves consumer response. AI is most effective when the value it creates is clearly legible to the consumer. This means that personalisation, recommendation, and service automation should be designed to feel supportive rather than invasive.

Second, transparency and proportionality in data practice are essential. The findings suggest that consumers are more willing to accept AI when data use feels legitimate and understandable. Marketers should therefore reduce opaque targeting practices, avoid overstating AI capability, and ensure that data capture remains proportionate to the benefit being delivered.

Third, firms should be cautious when deploying AI in emotionally expressive, hedonic, or morally sensitive communication. In such contexts, efficiency alone is unlikely to secure positive consumer response. Brands need to consider whether AI-generated or AI-disclosed communication may weaken authenticity, trust, or emotional connection.

Fourth, AI should be positioned as augmenting rather than replacing consumer agency. Systems that help consumers navigate choice, gather information, or receive service may be positively received, but systems that appear to narrow options, automate judgement too aggressively, or remove too much human presence may generate resistance.

\section{Limitations}
This study has several limitations. First, it is based entirely on secondary research and therefore depends on the scope, quality, and contextual diversity of existing studies. The dissertation does not provide new primary empirical evidence and cannot directly test the mechanisms proposed in the analysis.

Second, the literature review is interpretive in nature. Although the thematic framework provides coherence, the selection and synthesis of themes inevitably involve judgement. Different reviewers might emphasise somewhat different mechanisms, contextual conditions, or boundaries between themes.

Third, the field is developing rapidly, especially in relation to GenAI. As a result, some of the conclusions drawn here should be understood as provisional and open to revision as new empirical evidence emerges. A further limitation is that the dissertation title includes both AI and GenAI, while the available evidence is not evenly distributed across those two domains. Research on predictive AI remains more dominant in areas such as recommendation systems, personalised advertising, and data-driven targeting, whereas GenAI-specific evidence is much more concentrated in visible content generation, disclosure, and authorship-related contexts. This means that the four tensions identified in the study apply across AI more broadly, but the GenAI contribution is currently most developed in the social interaction and authenticity literature. Finally, the reviewed literature spans different platforms, markets, product categories, and cultural settings, which means that the general patterns identified in this study may not apply equally across all contexts.

\section{Future Research}
Future research should build on this dissertation in at least four directions. First, empirical studies are needed to test the mechanisms identified here, especially the roles of trust, perceived exploitation, perceived understanding, perceived control, and perceived human presence in shaping behavioural outcomes.

Second, future research should examine the conditions under which AI is more likely to be accepted or resisted. In particular, product type, involvement level, emotional intensity, cultural context, and disclosure strategy appear especially important and merit more systematic investigation.

Third, more research is needed on the distinction between source authenticity and emotional authenticity in GenAI marketing. This would help clarify when consumers object primarily to AI authorship itself and when they object more strongly to the emotional inadequacy of the resulting communication.

Fourth, future work should examine differences across consumer groups, markets, and levels of AI literacy. The literature suggests that not all consumers interpret AI in the same way, and a more differentiated understanding of these variations would strengthen both theory and practice.

Fifth, future research should extend GenAI inquiry beyond authorship and visible content effects. At present, GenAI is most often examined in relation to disclosure, authenticity, and AI-generated communication, while less is known about how GenAI may reshape the other three tensions through data capture, classification, and delegation. More work is therefore needed to determine whether GenAI introduces genuinely new perceptual mechanisms in those areas or intensifies dynamics already documented in predictive AI research.

\section{Final Remarks}
Artificial intelligence and generative AI are transforming the marketing landscape. While they offer significant opportunities for improving efficiency and personalisation, they also introduce new challenges for consumer perception.

This dissertation has argued that the central question is not whether AI in marketing is simply good or bad for consumers. The more important question is how consumers interpret the trade-offs AI creates, and under what conditions those trade-offs become acceptable or unacceptable. By showing that consumer perception is shaped by the tensions of being served or exploited, understood or misunderstood, empowered or replaced, and connected or alienated, the study provides a clearer framework for understanding the evolving relationship between AI, marketing, and consumer experience.
