\chapter{Methodology}

\section{Research Approach}
This dissertation adopts a qualitative research approach based on a literature review. Rather than collecting primary data, the study analyses existing academic literature on artificial intelligence (AI), generative artificial intelligence (GenAI), and consumer perception in marketing. This approach is appropriate because the research problem is conceptual and integrative in nature. The aim is not to test one causal relationship statistically, but to synthesise how the literature explains recurring patterns in consumer responses to AI-mediated marketing.

\section{Research Design}
The study uses a thematic literature review design. Thematic analysis was selected because the literature in this area is dispersed across different marketing applications, including personalised advertising, AI-generated content, recommendation systems, chatbots, and social media interactions. A thematic approach allows these studies to be brought together through recurring dimensions of consumer perception rather than through one single application area.

Importantly, this dissertation does not treat positive and negative findings as simple contradictions. Instead, it approaches the literature on the assumption that AI in marketing often generates conditional outcomes and recurring trade-offs. For example, the same AI application may improve relevance and convenience while also increasing privacy concern or reducing perceived authenticity. The methodological aim is therefore to identify not only the main themes in the literature, but also the mechanisms and contextual conditions through which those themes shape consumer responses.

\section{Data Collection}
The data for this study consists of peer-reviewed academic journal articles related to AI in marketing and consumer perception. Relevant literature was identified through academic databases such as Google Scholar, Scopus, and ScienceDirect. These databases were used because they provide broad coverage of marketing, consumer behaviour, and technology-focused research.

Keywords used for searching include:

\begin{itemize}
    \item Artificial intelligence in marketing
    \item Generative AI advertising
    \item Consumer perception AI
    \item AI-generated content
    \item AI trust, authenticity, and privacy
\end{itemize}

Additional searches combined these terms with related concepts such as recommendation systems, disclosure labels, chatbots, virtual influencers, consumer behaviour, and purchase intention. Backward and forward citation tracing was also used where necessary to identify influential studies repeatedly referenced in the field.

Because this dissertation is a thematic literature review rather than a formal systematic review, the search process was iterative rather than fully protocol-driven. Nevertheless, the narrowing process was made explicit. The final corpus used for thematic analysis comprised 26 peer-reviewed journal articles. In addition, two industry reports were retained only for contextual framing in the introduction and background chapter, not as part of the core academic evidence base. Sources were narrowed through staged screening for relevance at title, abstract, and full-text level, with conceptual fit to the research questions taking priority over exhaustive coverage.

\section{Data Selection Criteria}
To ensure relevance and quality, the following criteria were applied:

\begin{itemize}
    \item Peer-reviewed journal articles
    \item Primarily published between 2020 and 2026, with earlier studies included where they were conceptually foundational
    \item A clear focus on AI or generative AI in marketing or advertising contexts
    \item Direct discussion of consumer perception or closely related concepts such as trust, authenticity, privacy concern, relevance, usefulness, engagement, or purchase intention
    \item Sufficient conceptual or empirical relevance to the dissertation's research questions
\end{itemize}

Studies were excluded where they focused mainly on technical system performance, organisational adoption without a consumer dimension, or purely conceptual claims with limited relevance to marketing communication or consumer interpretation. The final body of literature was selected on the basis of relevance and analytical value rather than on a fixed numerical target alone.

In practice, the screening logic moved through three broad stages. First, an initial search and citation-chasing stage produced a broad candidate pool centred on AI, GenAI, consumer response, and digital marketing communication. Second, title and abstract screening removed studies that were clearly technical, organisational, or outside consumer-facing communication contexts. Third, full-text reading was used to retain only those studies that made a direct conceptual or empirical contribution to one or more of the dissertation's core themes. While an exact PRISMA-style record count was not maintained at the start of the project, the final included corpus and the inclusion logic are reported here to make the selection process more transparent.

\section{Thematic Analysis}
Thematic analysis was used to identify key themes across the selected literature and to explain how those themes operate. The process followed several stages:

\begin{enumerate}
    \item Familiarisation with the literature through repeated reading of selected articles
    \item Initial coding of recurring concepts, claims, and findings
    \item Grouping of codes into broader perceptual themes
    \item Comparison of studies to identify recurring trade-offs, mechanisms, and contextual conditions
    \item Refinement of themes into an integrated analytical framework
\end{enumerate}

The coding process was guided by recurrent concepts in the literature, including trust, authenticity, personalisation, privacy concerns, relevance, usefulness, emotional response, and behavioural intention. To structure these concepts more systematically, the dissertation uses Puntoni et al.'s experiential framework as its primary coding lens \cite{puntoni2021experiential}. Specifically, the literature was coded across four broad experience domains: data capture, classification, delegation, and social interaction. These domains were then translated into four consumer-perception tensions that recur across the reviewed studies:

\begin{itemize}
    \item Data capture: served versus exploited
    \item Classification: understood versus misunderstood
    \item Delegation: empowered versus replaced
    \item Social interaction: connected versus alienated
\end{itemize}

These tensions were not treated as isolated variables or formal statistical moderators. Instead, they were used as sensitising themes that helped interpret how consumer responses to AI are produced across different contexts. A fuller coding table showing the relationship between theme groups, key constructs, supporting studies, and interpretive tensions is provided in the Appendix.

Trust requires a further methodological clarification because its role changes slightly between the literature review and the findings chapter. During the initial review stage, trust was retained as a visible thematic heading because it appears repeatedly in prior studies as an explicit construct and often functions as a reported predictor, mediator, or outcome. During the later stage of thematic refinement, however, trust was repositioned as a cross-cutting mechanism because the coding process showed that it emerged across all four experience domains rather than functioning as one self-contained category. This shift was therefore not an ad hoc contradiction in the analysis, but the result of moving from article-level topic coding to cross-study thematic synthesis.

The final analytical framework focuses on three levels of interpretation. First, it identifies the main dimensions through which AI shapes consumer perception. Second, it examines the mechanisms that explain why these effects occur, such as perceived relevance, emotional sincerity, opacity, intrusiveness, or loss of autonomy. Third, it considers the conditions under which responses become more positive or more negative, including AI visibility, message type, product context, and the extent of data use.

To avoid conceptual ambiguity, this dissertation distinguishes between three related but non-identical levels. The first is interpretive appraisal, which refers to how consumers cognitively judge AI as useful, credible, appropriate, intrusive, or fair. The second is affective response, which refers to feelings such as trust, unease, alienation, curiosity, or moral discomfort. The third is behavioural outcome, which refers to consequences such as engagement, ad avoidance, purchase intention, reliance, or resistance. The term consumer perception is therefore used as an umbrella concept that refers most directly to interpretive appraisal and affective response, while behavioural outcomes are treated as consequences shaped by those evaluations rather than as perception itself.

\section{Limitations}
This study has several limitations. First, it relies on secondary data and is therefore dependent on the scope, quality, and contexts of existing published studies. Second, literature selection in a thematic review inevitably involves interpretive judgement, which may shape how themes and mechanisms are prioritised. Third, the field is rapidly developing, particularly in relation to GenAI, meaning that current conclusions should be understood as provisional rather than final. A further limitation is that the evidence base is not evenly distributed across AI types: studies of predictive AI remain more prominent in areas such as personalisation, recommendation, and data-driven targeting, whereas GenAI evidence is more concentrated in visible content creation, disclosure, and authorship-related contexts. Finally, because the dissertation synthesises studies conducted across different markets, platforms, and product categories, some findings may not generalise equally across all consumer contexts.

\section{Summary}
This chapter has outlined the qualitative literature review approach, the selection of academic sources, and the thematic analysis process used in the study. It has also clarified that the purpose of the analysis is not simply to catalogue positive and negative findings, but to explain the mechanisms and conditions through which AI influences consumer perception in marketing. The next chapter presents the findings and discussion developed from this analytical framework.
