\chapter{Literature Review}

\section{Introduction}
This chapter reviews the literature on artificial intelligence (AI), generative artificial intelligence (GenAI), and consumer perception in marketing. Its main purpose is to establish the conceptual foundation for this dissertation and to identify the recurring themes that shape how consumers respond to AI-mediated marketing activity. While AI has long been used to support segmentation, targeting, forecasting, and automation, GenAI has made AI far more visible within marketing communication itself. This shift matters because consumers are no longer responding only to products, brands, or platforms. They are also responding to the role AI plays in creating, personalising, and delivering marketing messages.

Although the literature on AI in marketing has expanded rapidly, it remains somewhat fragmented. Some studies treat AI primarily as a strategic or operational capability that improves efficiency and decision-making, while others focus on consumer responses to AI-generated content, disclosure labels, recommendation systems, or personalised advertising \cite{davenport2020future,haleem2022ai,kumar2024aipowered,chan2025genai}. As a result, the field contains a wide range of useful insights, but these are often spread across different applications and theoretical traditions. This matters because the same technology may be evaluated very differently depending on whether it is used to improve convenience, automate service, or produce emotionally expressive marketing communication. This chapter therefore adopts a consumer-centred perspective and synthesises the literature around the themes most relevant to this dissertation: trust, authenticity, personalisation, privacy concerns, and behavioural outcomes.

\section{Artificial Intelligence in Marketing}
AI is now deeply embedded in contemporary marketing practice. Earlier discussions of AI in marketing often focused on its functional value, particularly its ability to analyse large datasets, predict customer behaviour, automate routine decisions, and improve targeting accuracy \cite{haleem2022ai,jain2024twodecades}. In this earlier phase, AI largely operated in the background as a decision-support tool. Its value lay in efficiency, optimisation, and scale rather than in direct visibility to consumers.

More recent work suggests that AI should be understood not merely as a technical tool, but as a force that is reshaping the logic of marketing itself. Davenport et al. argue that AI is changing the future of marketing by altering how firms gather customer insights, design interactions, automate communication, and make strategic decisions \cite{davenport2020future}. Kumar et al. similarly show that AI now extends across multiple marketing functions, including customer engagement, campaign design, content delivery, and decision automation \cite{kumar2024aipowered}. In parallel, Mohd Amin et al. suggest that the adoption of AI in marketing strategies is transforming business landscapes by making predictive analytics and intelligent personalisation central to value creation \cite{mohdamin2025adoption}.

This broader view is important for the present dissertation because it highlights that AI does more than make marketing more efficient. It also changes how firms relate to consumers and how consumers interpret those interactions. Earlier studies are generally more optimistic when AI remains in the background as an analytical or service-support tool. By contrast, more recent work becomes more cautious when AI is visible in communication itself, because consumers begin to judge not only the outcome of marketing but also the source and meaning of the message. Once AI becomes visible within advertising, service encounters, and branded communication, consumer perception becomes a central concern rather than a secondary outcome.

\section{From Predictive AI to Generative AI}
The emergence of GenAI marks a significant shift in the literature. Hermann and Puntoni distinguish between earlier predictive forms of AI, which mainly support decisions behind the scenes, and generative systems that actively produce text, images, and other content experienced directly by consumers \cite{hermann2024consumer}. This distinction is particularly important in marketing because it moves AI from an invisible analytical infrastructure to a visible participant in communication.

Chan and Choi argue that GenAI is rapidly becoming embedded across marketing practice, including advertising, service design, creative production, and innovation processes \cite{chan2025genai}. Their work is useful because it shows that GenAI should not be treated simply as an extension of traditional AI adoption. Instead, it raises new questions about content ownership, disclosure, ethics, and consumer acceptance. In a similar vein, Ubal et al. contend that consumer responses to AI depend not only on the technical capabilities of the system, but also on the psychological and contextual cues that shape whether AI is interpreted as useful, appropriate, competent, or socially acceptable \cite{ubal2024framework}. A key implication here is that consumer reactions are often more positive when AI enhances speed, relevance, or accessibility, but more negative when it appears to replace human creativity or emotional judgement. For a dissertation concerned with perception, this shift from predictive AI to GenAI is especially significant because it introduces stronger tensions between efficiency, creativity, trust, and authenticity.

\section{Consumer Perception as a Conceptual Lens}
Consumer perception provides a useful lens through which to examine AI in marketing because it directs attention away from technical capability alone and towards interpretation, evaluation, and response. In this context, the central issue is not simply what AI can do, but how consumers make sense of its presence in the marketplace and how that interpretation shapes subsequent attitudes and behaviours.

Puntoni et al. offer one of the most useful conceptual foundations for this line of inquiry. They argue that consumer experiences with AI are shaped by recurring tensions linked to data capture, classification, delegation, and social interaction \cite{puntoni2021experiential}. These experiences may generate convenience and perceived usefulness, but they may also produce feelings of exploitation, misunderstanding, replacement, or alienation. This perspective is particularly valuable because it explains why consumer responses to AI are often mixed rather than consistently positive or negative. It also helps connect a wide range of empirical findings under a broader experiential framework.

Related review work reinforces this multidimensional understanding of consumer perception. Ubal et al. propose that AI-consumer interaction is influenced by psychological, contextual, and relational factors rather than by one single evaluative criterion \cite{ubal2024framework}. Similarly, Hermann and Puntoni suggest that the move towards GenAI intensifies interpretive complexity because consumers are increasingly confronted with AI as a creator rather than merely a predictor \cite{hermann2024consumer}. Taken together, this literature suggests that trust, authenticity, personalisation, privacy, and behavioural intention should be treated as interconnected dimensions of consumer perception rather than isolated variables.

\section{Key Themes in AI Marketing and Consumer Perception}

\subsection{Trust and AI-Enabled Customer Experience}
Trust is one of the most consistently discussed themes in the AI marketing literature. It generally refers to the extent to which consumers perceive AI-enabled systems and AI-mediated messages as reliable, competent, and aligned with their interests. Trust matters because it often mediates the relationship between technological capability and favourable consumer response.

Ameen et al. make this point clearly in their study of AI-enabled customer experience. They show that trust plays a central role in shaping how consumers evaluate AI-supported service interactions, especially where convenience, personalisation, and perceived service quality are involved \cite{ameen2021customer}. Their work is especially useful for this dissertation because it demonstrates that AI-driven experiences are not evaluated solely on efficiency grounds. Consumers also care about whether those experiences feel dependable, understandable, and worth the trade-offs involved.

Other studies reinforce the importance of trust in more specific marketing contexts. An and Ngo find that trust is strengthened when AI-powered personalised advertising is perceived as relevant and useful, which in turn supports purchase intention \cite{an2025personalized}. In contrast, Israfilzade shows that consumer trust can weaken when advertising is known to be AI-generated rather than human-created, especially where consumers interpret AI authorship as reducing credibility or sincerity \cite{israfilzade2025advertising}. In social media settings, Srivastava and Gurme similarly find that trust remains a key pathway through which personalisation influences consumer behaviour, although this effect is weakened when privacy concerns become salient \cite{srivastava2026social}. What emerges across these studies is a clear contextual difference: AI tends to receive more positive trust evaluations when it supports relevance and convenience behind the scenes, but more negative evaluations when it becomes highly visible in the message itself or when data use feels opaque. Overall, the literature suggests that trust is not an automatic outcome of technological sophistication; it depends on whether consumers see AI as useful, legitimate, and socially acceptable.

\subsection{Authenticity, Emotion, and AI-Generated Content}
Authenticity has become especially prominent in the GenAI literature because AI-generated content raises questions about originality, sincerity, and human presence. This issue is particularly important in marketing communication, where consumers often respond not only to informational content but also to perceived emotional intention and symbolic meaning.

Arango, Singaraju, and Niininen offer strong evidence that AI-generated content can alter emotional response in meaningful ways. In their study of charitable giving advertisements, AI-generated ads reduced perceived authenticity and emotional connection, which in turn affected empathy and donation intention \cite{arango2023charitable}. This is highly relevant to the present dissertation because it shows that AI-generated content may influence consumers even when the informational message remains broadly similar. What changes is the perceived genuineness of the communication and the emotional bond it creates.

Kirk and Givi make a related argument in their discussion of the ``AI-authorship effect''. They show that consumers respond less favourably to emotionally expressive marketing communication when they believe it was produced by AI rather than by a human, largely because perceived authenticity declines \cite{kirk2025aiauthorship}. Shi and Jiang similarly demonstrate that AI disclosure labels can generate a mixed response: they may increase novelty, but they can also lower authenticity and weaken consumer attitudes or purchase intention \cite{shi2026disclosure}. These studies are broadly consistent, but they also suggest an important nuance. Negative responses become strongest in contexts where consumers expect empathy, sincerity, or human creativity, whereas novelty may partly offset those concerns in lower-stakes or more experimental settings. Taken together, these studies suggest that authenticity should be treated as a central mechanism in consumer evaluations of GenAI marketing content rather than as a peripheral aesthetic concern.

\subsection{Personalisation, Relevance, and Perceived Usefulness}
One of the most widely cited benefits of AI in marketing is its capacity to deliver personalised experiences. AI can tailor recommendations, advertisements, and service encounters in ways that increase relevance, reduce search effort, and potentially improve customer satisfaction \cite{haleem2022ai,kumar2024aipowered}. This makes personalisation a central justification for AI adoption from a managerial perspective.

However, the literature shows that personalisation is not inherently persuasive. Its positive effect depends on how consumers interpret it. Ameen et al. suggest that AI-enabled customer experience becomes more favourable when consumers perceive benefits such as convenience and improved service quality, but this evaluation is still shaped by trust and perceived sacrifice \cite{ameen2021customer}. Similarly, An and Ngo find that the effect of AI-powered personalised advertising on purchase intention is strengthened when personalisation is understood as relevant and useful rather than merely technologically advanced \cite{an2025personalized}. This implies that the same mechanism can produce both positive and negative responses: personalisation is welcomed when it reduces effort and improves fit, but it becomes problematic when it feels excessive, mechanistic, or too closely tied to data surveillance. These studies suggest that personalisation works best when it feels helpful and proportionate, not when it appears intrusive or overly automated.

\subsection{Privacy Concerns, Transparency, and Perceived Intrusiveness}
Privacy concerns remain one of the most persistent tensions in the literature on AI marketing. Since many AI applications rely on the collection and interpretation of behavioural and personal data, consumers may feel uneasy about tracking, profiling, or opaque data practices. This issue becomes especially important when personalisation is perceived to cross the boundary from relevance into surveillance.

Boerman et al., although writing in the context of online behavioural advertising rather than GenAI, provide an important foundation for understanding this issue. Their review highlights how personalisation, transparency, disclosure, and perceived intrusiveness are closely linked, and it shows that consumer responses to targeted advertising are shaped by whether data use feels visible, fair, and controllable \cite{boerman2017oba}. This earlier literature remains highly relevant because AI-driven personalisation intensifies many of the same tensions.

More recent AI-focused studies confirm that privacy remains a major barrier to consumer acceptance. Davenport et al. identify privacy and bias as structural challenges in the future development of AI marketing \cite{davenport2020future}. Kumar et al. and Mohd Amin et al. likewise show that concerns around data use and ethical governance continue to shape the adoption and evaluation of AI systems in marketing \cite{kumar2024aipowered,mohdamin2025adoption}. Empirical evidence from social media contexts also supports this view, with Srivastava and Gurme finding that privacy concerns weaken favourable behavioural responses even where personalisation and trust are otherwise present \cite{srivastava2026social}. Compared with the more positive literature on relevance and convenience, this work shows that favourable responses to AI are conditional rather than stable. Privacy, then, is not just an ethical issue in the abstract; it is a perceptual barrier that can directly undermine the effectiveness of AI-enabled marketing practice.

\subsection{Behavioural Outcomes: Purchase Intention, Engagement, and Donation Intention}
Many studies ultimately connect consumer perception to behavioural outcomes such as purchase intention, engagement, recommendation, and donation intention. This is significant because it demonstrates that perceptual themes are not merely abstract attitudes. They have practical consequences for market behaviour.

Existing evidence suggests that AI can improve behavioural outcomes when consumers interpret its use as relevant, useful, and supportive of their goals. An and Ngo, for example, show that relevance, usefulness, and trust positively contribute to purchase intention in AI-powered personalised advertising \cite{an2025personalized}. However, these effects are not stable across all contexts. Arango et al. show that reduced authenticity and emotional connection can lower donation intention in charitable advertising \cite{arango2023charitable}. Research on GenAI visual content similarly suggests that behavioural responses vary depending on context, involvement level, and the reasons given for AI use. Belanche et al. find that real images are often preferred in high-involvement and hedonic service contexts, while Zhang and Hur show that disclosure of motives such as privacy protection, visual appeal, or cost efficiency can alter consumer attitudes towards AI-generated advertising images \cite{belanche2025images,zhang2025images}. The pattern here is not contradictory so much as conditional: AI tends to support positive behavioural outcomes when it delivers practical value without challenging expectations of sincerity, but it becomes less effective when the product category, message type, or emotional context makes human authenticity more important. These findings indicate that behavioural outcomes are shaped by the interaction of several perceptual mechanisms rather than by AI use alone.

\section{Synthesis and Research Gap}
Taken together, the literature suggests that AI in marketing is best understood through a series of interrelated perceptual tensions. AI can improve convenience, relevance, and efficiency, yet it may also weaken authenticity, reduce emotional connection, and intensify privacy concerns. These tensions become more visible in GenAI contexts because AI is no longer only supporting decisions behind the scenes. It is increasingly involved in producing the very content consumers see and evaluate.

Several gaps remain despite the rapid growth of this field. First, the literature is still fragmented across different applications such as service automation, personalised advertising, disclosure labels, social media engagement, and AI-generated visual or textual content. Second, many studies focus on one variable at a time, such as trust, authenticity, or privacy, without fully explaining how these factors interact in shaping consumer responses. Third, while some empirical studies report positive outcomes for AI when usefulness and convenience are salient, others report negative outcomes when authenticity, emotional sensitivity, or privacy become more important. These differences are often context-specific, but the literature does not yet integrate them clearly enough into one consumer-centred explanation. Finally, broad conceptual discussions of AI and GenAI in marketing often sit separately from narrower empirical studies, making it difficult to construct an integrated account of when AI is perceived positively and when it is resisted \cite{chan2025genai,puntoni2021experiential,ubal2024framework}. This dissertation addresses these gaps by conducting a thematic literature review focused on the key dimensions through which consumers interpret AI and GenAI in marketing environments.

\section{Summary}
This chapter has reviewed the literature on AI and GenAI in marketing from a consumer perception perspective. It has shown that AI in marketing cannot be assessed only in terms of efficiency or innovation. Consumer responses are shaped by a more complex set of perceptual dimensions, especially trust, authenticity, personalisation, privacy concerns, and behavioural intention. These themes provide the conceptual basis for the methodology and thematic analysis developed in the following chapters.